---
title: "Lecture 1a: Linear Congruential Generator"
subtitle: "Simulations in Statistics (52001)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

## Overview

A **Linear Congruential Generator (LCG)** produces a sequence of pseudo-random integers via the recursion:

$$X_{n+1} = (a \cdot X_n + b) \pmod{m}$$

where:

- $X_0$ is the **seed** (initial value)
- $a$ is the **multiplier**
- $b$ is the **increment**
- $m$ is the **modulus**

The uniform pseudo-random variable at step $n$ is:

$$U_n = \frac{X_n + 1}{m + 1}$$

The $+1$ shift maps $X_n \in \{0, \ldots, m-1\}$ onto $(0, 1)$, avoiding the degenerate endpoints.

## R Implementation

```{r lcg-function}
Part1a <- function(n, a, b, m, seed) {
  x_current <- seed

  for (step_index in seq_len(n)) {
    x_current <- (a * x_current + b) %% m
  }

  u_n <- (x_current + 1) / (m + 1)
  return(u_n)
}
```

## Parameters

```{r parameters}
n_steps    <- 79
multiplier <- 19
increment  <- 15
modulus    <- 64
seed_value <- 17

cat("Parameters:\n")
cat("  n    =", n_steps,    "\n")
cat("  a    =", multiplier, "\n")
cat("  b    =", increment,  "\n")
cat("  m    =", modulus,    "\n")
cat("  seed =", seed_value, "\n")
```

## Computing $U_n$

```{r compute-u-n}
u_n <- Part1a(n = n_steps, a = multiplier, b = increment, m = modulus, seed = seed_value)
cat("U_n =", u_n, "\n")
```

## Closed-Form Derivation

Rather than iterating $n$ times, the closed-form formula jumps directly to $X_n$:

$$X_n = \left( a^n X_0 + b \cdot \frac{a^n - 1}{a - 1} \right) \pmod{m}$$

which simplifies to:

$$X_n = \left( \left(\frac{b}{a-1} + X_0\right) \cdot a^n - \frac{b}{a-1} \right) \pmod{m}$$

```{r closed-form}
coefficient <- as.integer(increment / (multiplier - 1) + seed_value)
offset      <- as.integer(increment / (multiplier - 1))

a_pow_n_mod_m <- as.integer(Reduce(
  function(acc, .x) (acc * multiplier) %% modulus,
  seq_len(n_steps),
  init = 1
))

x_n       <- (coefficient * a_pow_n_mod_m - offset) %% modulus
u_n_exact <- (x_n + 1) / (modulus + 1)

cat("a^n mod m =", a_pow_n_mod_m, "\n")
cat("X_n       =", x_n,           "\n")
cat("U_n       =", u_n_exact,     "\n")
```

## Final Answer

$$U_n = \frac{X_n + 1}{m + 1}$$
